---
title: "Give your ggplots a Taylor Swift theme"
output: rmarkdown::html_vignette
description: |
  Make your ggplots *gorgeous*.
vignette: >
  %\VignetteIndexEntry{Give your ggplots a Taylor Swift theme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(taylor)
```

In addition to data sets, taylor also includes a few helper functions for easily making plots created with [ggplot2](https://ggplot2.tidyverse.org) have a Taylor Swift theme. For this vignette, I'm assuming you are already familiar with ggplot2 and can make basic plots. 

## ggplot2 scales

You can add album-themed color palettes to your plots using the `scale_color_taylor_d()` and `scale_fill_taylor_d()` for discrete scales, and using `scale_color_taylor_c()` and `scale_fill_taylor_c()` for continuous scales. We'll start will discrete scales.

### Discrete scales

First, let's make a bar graph showing the valence of each song on *evermore*.

```{r warning = FALSE, message = FALSE}
library(taylor)
library(ggplot2)

evermore <- tswift_album_songs[tswift_album_songs$album_name == "evermore", ]
evermore$track_name <- factor(evermore$track_name, levels = evermore$track_name)

p <- ggplot(evermore, aes(x = valence, y = track_name, fill = track_name)) +
  geom_col(show.legend = FALSE) +
  expand_limits(x = c(0, 1)) +
  labs(y = NULL) +
  theme_minimal()
p
```

We can then add some *evermore*-inspired colors using `scale_fill_taylor_d()`.

```{r}
p + scale_fill_taylor_d(album = "evermore")
```

The `album` argument can be changed to use a different Taylor-inspired palette. For example, we can switch to *Speak Now* using `album = "Speak Now"`.

```{r}
p + scale_fill_taylor_d(album = "Speak Now")
```


### Continuous scales



## Color palettes

